---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Robert Ford
thanks: "Code and data are available at: https://github.com/Ford-Robert/STA304_City-Of-Toronto-Data.git"
date: 27/09/2024
date-format: 27 September 2024
abstract: "This study analyzes ten years of delay data from the Toronto Transit Commission (TTC), focusing on subways, streetcars, and buses to assess changes in delay times of these services. The analysis found that buses experience the highest number of delays and the longest total and average delay times, while subway delays are resolved much faster on average. The analysis also delves into the specific challenges each service faces, analyzing which incidents they are most likely to experience and how long those incidents take to resolve. These findings provide data-driven insights for improving TTC operations and uncovering the specific challenges faced by each transportation mode."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(dplyr)
library(knitr)
library(ggplot2)
library(scales)
library(gridExtra)
library(patchwork)
```

```{r}
#| message: false
#| echo: false
#| warning: false
data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))

```



# Introduction{#sec-introduction}
Effective public transportation is vital to any large dense urban area. It has the capacity to significantly reduce traffic congestion [REF] and improve transportation times for all residents of a city. Public transportation also provides affordable transportation options allowing for a more equitable society, in which everyone has access to opportunities across a city [REF]. Expanding TTC services, and ensuring reliability, is the best option for taking cars off the road which would directly reduce both emissions [REF] and car related deaths within the city [REF]. 

Despite the importance of the TTC, public opinion in Toronto suggests that the TTC is failing to meet the needs of its riders. Complaints about aging infrastructure, frequent delays, and most glaringly safety concerns are widely shared among Torontonians [REF]. Another point of criticism is that the TTC, especially the subway, has failed to expand much in recent years, despite Toronto growing significantly in population. 

Considering these concerns this paper analyzes TTC delay data from 2014 to 2024, to assess whether its service have truly deteriorated over the past decade. By examining delays across subways, streetcars, and buses, this study will first compare these services with each other to explore which suffers the most from high delay times and large numbers of incidents. Then this study aims to identify the specific challenges each service faces. Finally, this paper will explore how delay times and incident rates are distributed over days of the week, and how these metrics have trended over the last ten years. 

These findings offer insights for policymakers, transit authorities, and the public. By taking a deep dive into the publicly available data, this analysis provides a foundation for understanding TTC delays across buses, streetcars, and subways. The paper includes an exploration of the data by constructing easy-to-read graphs, designed to highlight the areas of greatest concern within the TTC. 


# Data {#sec-data}

Delay Data detailing the operations of the TTC was collected and cleaned from the Open Data Toronto [REF] website. This data spans about 10 years, from 2014 to 2024. Three separate datasets were used, Delay data related to the Subway[REF], Streetcar[REF] and Bus[REF]. The original datasets are separated into a variety of formats depending on which year you are examining. The way in which variable names and the data stored changes over time. To further complicate the collection process, the information found in each dataset was different, though it mostly represented the same concepts. Data that was too difficult to integrate into one combined dataset that incorporated all ten years, and all three modes of transport was excluded. The final dataset contains 1,042,564 observations of 6 variables; 

  -Date: The Year, Month and Day that a delay occurred 

  -Day: What day of the week that a delay occurred 

  -Vehicle: What mode of transport did the delay occur on 

  -Location: Where did the delay occur. For Subways Location is the nearest station. For buses and                 streetcars Location is the nearest major intersection or passenger stop. 

  -Incident: A description of what caused the delay 

  -Delay: How long in minutes the delay lasted 

```{r data-table}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: First five rows of the Final Dataset
#| fig-align: center

kable(head(data, 5), caption = "Sample of the Data")
```
 

In the cleaning phase, all data entries that had an incomplete (NA) value for the Delay variable were removed. Because if a delay occurred but we do not how impactful that delay was, then it does not contribute to the analysis. Data that was clearly erroneous was removed, like negative delay times and extremely high delay times. Furthermore, all observations whose delay time exceeded 998 minutes (999 minutes was the code for a cancellation between 2015 and 2019) were removed. 


## Delay of Vehicles {#subsec1-data}

First this paper will assess which Mode of Transport suffers the most from delays. From @fig-incidents-mode, we observe that buses suffer the highest number of delays and are delayed for the longest period. Interestingly, though the subways suffer from more Incidents than the streetcar, its total time lost to delays is significantly less. This would indicate that individual Subway incidents are resolved faster than Streetcar incidents on average. 
```{r fig-incidents-mode}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: (left) Total Number of Incidents (right) Total Amount of Delay Time
#| fig-align: center

plot1 <- ggplot(data, aes(x = vehicle)) +
  geom_bar(fill = "steelblue") +
  scale_y_continuous(labels = comma) +
  labs(title = "Number of Incidents by Vehicle Type", x = "Vehicle Type", y = "Number of Incidents") +
  theme_minimal()

cumulative_delay <- data %>%
  group_by(vehicle) %>%
  summarise(total_delay = sum(Delay, na.rm = TRUE))


plot2 <- ggplot(cumulative_delay, aes(x = vehicle, y = total_delay)) +
  geom_bar(stat = "identity", fill = "tomato") +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative Delay Time by Vehicle", x = "Vehicle Type", y = "Total Delay (Minutes)") +
  theme_minimal()

grid.arrange(plot1, plot2, ncol = 2)

```


```{r fig-ave-delay}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time by Vehicle Type
#| fig-align: center

average_delay <- data %>%
  group_by(vehicle) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE))

ggplot(average_delay, aes(x = vehicle, y = avg_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Delay Time by Vehicle Type", x = "Vehicle Type", y = "Average Delay (Minutes)") +
  theme_minimal()

```

@fig-ave-delay would support this claim, as on average Streetcar delays take 13 minutes to resolve whereas Subway delays only take about 2.30 minutes to resolve. Unfortunately, Buses not only have the most delays, but its delays on average take about 20 minutes to resolve.


## Causes of Delays {#subsec2-data}
To discover why Bus delays are so much worse than Subway and Streetcar delays, we need to observer the differences in how long every type of incident takes to resolve, as well as what type of incident each mode of transport suffers from the most. This will also provide crucial insight into which Incidents are the most problematic, then the TTC may focus their efforts on the most disruptive incidents.  

```{r fig-total-delay}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time for Buses by, Incident Type
#| fig-align: center


# Step 1: Calculate total delay time for each incident type
data_incident_total_delay <- data %>%
  group_by(Incident) %>%
  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop') %>%
  mutate(Incident = fct_reorder(Incident, total_delay, .desc = TRUE))  # Reorder by total_delay

# Step 2: Create the bar chart for total delay time by incident type
ggplot(data_incident_total_delay, aes(x = Incident, y = total_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Delay Time by Incident Type", 
       x = "Incident Type", y = "Total Delay Time (Minutes)") +
  scale_y_continuous(labels = comma) +  # Format y-axis labels to be whole numbers with commas
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))

```
From @fig-total-delay we can see that Diversions make up the bulk of the total time lost to delays over the last 10 years. This could be due to the sheer number of Diversions that happen, or each Diversion causes a large delay. By investigating @fig-delay-incident we observe that Diversions on average take a long time to resolve.

```{r fig-delay-incident}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time by Incident Type
#| fig-align: center

data_incident_avg_delay <- data %>%
  group_by(Incident) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay

# Create the bar chart
ggplot(data_incident_avg_delay, aes(x = Incident, y = avg_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
  labs(title = "Average Delay Time by Incident Type", 
       x = "Incident Type", y = "Average Delay (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))


data_incident_by_delay <- data %>%
  group_by(Incident, vehicle) %>%
  summarise(incident_count = n(), avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
  group_by(Incident) %>%
  mutate(percentage = (incident_count / sum(incident_count)) * 100) %>%
  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))

```
Next, we will break down average delay time by each mode of transport, as this will give us a better understanding of incidents are the most disruptive for each mode of transport. In @fig-delay-bus we observe that buses are most impacted by Held by, Overhead, and Diversions. To most improve Bus services the TTC should focus on reducing the average delay times of these Incidents. @fig-delay-streetcar displays that the Streetcar is plagued by the same high-cost average delay time incidents. However, the spread for Streetcar is broader so targeting these three incidents would not be as impactful as it is for buses. However, @fig-delay-subway shows part of why the subway suffers so much less from delays. The incidents the subway suffers from take only 4 minutes on average to resolve, in the worst case.

```{r fig-delay-bus}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time for Buses by, Incident Type
#| fig-align: center


# Prepare the data: calculate the average delay time for each incident type and vehicle
data_incident_avg_delay <- data %>%
  group_by(vehicle, Incident) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')

# Function to create a bar chart for a specific vehicle
create_bar_chart <- function(vehicle_type) {
  # Filter data for the specific vehicle type and reorder incidents by average delay
  vehicle_data <- data_incident_avg_delay %>%
    filter(vehicle == vehicle_type) %>%
    mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
  
  # Create the bar chart with uniform color and spread the x-axis labels
  ggplot(vehicle_data, aes(x = Incident, y = avg_delay)) +
    geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
    labs(title = paste("Average Delay by Incident Type for", vehicle_type),
         x = "Incident Type", y = "Average Delay (Minutes)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))  # Rotate labels for readability
}

# Generate the separate bar charts for each vehicle type
bus_bar_chart <- create_bar_chart("Bus")
bus_bar_chart

```
```{r fig-delay-streetcar}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time for Streetcars, by Incident Type
#| fig-align: center
streetcar_bar_chart <- create_bar_chart("Streetcar")
streetcar_bar_chart
```
```{r fig-delay-subway}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time for Subways, by Incident Type
#| fig-align: center

subway_bar_chart <- create_bar_chart("Subway")
subway_bar_chart

```

## Delays Over Time {#subsec3-data}

To further investigate the data, we will investigate when incidents are likely to occur. From @fig-delay-days we observe that the average delay time of incidents is about 2 minutes longer on the weekends. This could be because there are less TTC staff on the weekends, so incidents take longer to be delt with. Across months, @fig-delay-months, there is little difference in the average delay time. This is surprising, as in the winter delays could be worse as weather conditions worsen. This may be a testament to the TTC’s ability to deal with Toronto’s harsh winters. 
```{r fig-delay-days}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time by Day
#| fig-align: center
# Step 1: Extract the day of the week from the Date column and calculate average delay time by day of the week
data_avg_delay_by_day <- data %>%
  mutate(day_of_week = wday(Date, label = TRUE, abbr = FALSE)) %>%  # Extract day of the week from Date
  group_by(day_of_week) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')

# Step 2: Create the bar chart for average delay by day of the week
ggplot(data_avg_delay_by_day, aes(x = day_of_week, y = avg_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Vertical bars for average delay
  labs(title = "Average Delay Time by Day of the Week", 
       x = "Day of the Week", y = "Average Delay Time (Minutes)") +
  theme_minimal()

```


```{r fig-delay-months}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Amount of Delay Time by Month
#| fig-align: center

# Step 1: Extract the month from the Date column and calculate average delay time by month
data_avg_delay_by_month <- data %>%
  mutate(month = month(Date, label = TRUE, abbr = FALSE)) %>%  # Extract month from Date and get full month names
  group_by(month) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')

# Step 2: Reorder the months to display them in chronological order
data_avg_delay_by_month <- data_avg_delay_by_month %>%
  mutate(month = factor(month, levels = month.name))  # Ensure months are ordered chronologically

# Step 3: Create the bar chart with switched axes
ggplot(data_avg_delay_by_month, aes(x = month, y = avg_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Vertical bars with average delay
  labs(title = "Average Delay Time by Month", 
       x = "Month", y = "Average Delay Time (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))  # Rotate x-axis labels for readability
```


Finally, to get a wider view of how incidents have changed over the past 10 years we look to @fig-over-time and @fig-total-delay In @fig-over-time we see that the number of bus incidents decreased dramatically between 2014 and 2021. Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in @fig-total-delay. This may be due to the TTC changing how they classify incidents, so the data is entered differently but the amount of delay time remains stable.
```{r fig-over-time}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Number of Incidents Over Time
#| fig-align: center


data_inc_num <- data %>%
  group_by(month = floor_date(Date, "month"), vehicle) %>%
  summarise(incident_count = n())

# Create the line graph
ggplot(data_inc_num, aes(x = month, y = incident_count, color = vehicle, group = vehicle)) +
  geom_line(size = 1) +
  labs(title = "Number of Incidents Over Time by Vehicle", x = "Time", y = "Number of Incidents", color = "Vehicle Type") +
  theme_minimal()

```

```{r fig-total-delay}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Total Delay Time from 2014 to 2024
#| fig-align: center
# Step 1: Convert the Date column to Date format (if it's not already) and group by month and vehicle type
data_total_delay_time <- data %>%
  mutate(Date = as.Date(Date)) %>%  # Ensure Date is in Date format
  group_by(month = floor_date(Date, "month"), vehicle) %>%  # Group by month and vehicle type
  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop')  # Summarize total delay

# Step 2: Create the line chart with 3 lines for each mode of transport
ggplot(data_total_delay_time, aes(x = month, y = total_delay, color = vehicle, group = vehicle)) +
  geom_line(size = 1) +  # Create the line chart with separate lines for each vehicle
  labs(title = "Total Incident Time by Mode of Transport (2014-2024)", 
       x = "Time", y = "Total Incident Time (Minutes)", color = "Vehicle Type") +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Format the x-axis with yearly labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```
@fig-total-delay suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop. This drop could be related to the Covid-19 pandemic for several reasons. Firstly, ridership dropped dramatically during the pandemic therefore fewer buses were scheduled and thus fewer Diversions, Overhead and Collisions. There is a modest drop in Streetcar Delay times as well which would support this narrative. Secondly, the city made public transportation free during the pandemic, so delays due to Investigations would fall. In @fig-ave-delaytime observe that average delay time did not noticeably drop during the pandemic. This lends further support to these claims, as average Incident delay time remained the same, however far fewer incidents were occurring as the system was constricted. Furthermore, Fig10 indicates that over the last 10 years the average delay time per incident has been increasing slightly. This could be an indication that the aging fleet of TTC vehicles requires more time to resolve certain incidents.

```{r fig-ave-delaytime}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Delay Time from 2014 to 2024
#| fig-align: center

data_monthly_delay <- data %>%
  group_by(month = floor_date(Date, "month"), vehicle) %>%
  summarise(average_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')

# Create the line graph for monthly average delay times
ggplot(data_monthly_delay, aes(x = month, y = average_delay, color = vehicle, group = vehicle)) +
  geom_line(size = 1) +
  labs(title = "Average Delay Time Over Time by Vehicle (Monthly)", 
       x = "Time (Month)", y = "Average Delay Time (Minutes)", color = "Vehicle Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```



# Discussion {#sec-discussion}

The results show that buses experience the highest number of delays and the longest total and average delay times. Incidents that have long average delay times such as diversions, mechanical issues, and “held by” incidents contribute most to the total delays buses suffer from. In contrast, subway delays are resolved more quickly on average, which may be attributed to less exposure to external factors like traffic, and a more controlled environment allowing for faster repair or replacement. 

These findings have important implications for the TTC and its efforts to improve service reliability. For example, reducing the frequency and duration of diversions and mechanical issues could have a disproportionate effect on improving bus services. Expanding the Subway would be a great strategy as subway services are inherently more reliable, so maximizing ridership there would be beneficial.

To build on these findings, further exploration focusing on location data could provide deeper insights into delay patterns across the city. Creating a heat map of Toronto that visualizes delay times would identify geographic hotspots where delays are most frequent or severe. This spatial analysis could help TTC officials focus their efforts on certain neighborhoods, intersections, or corridors that are more prone to delays. 


\newpage


# References


::: {#refs}
:::